---
title: "HIV Epidemiology – Children & Adolescents"
author: "Akanshya Dash"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(here)
  library(readxl)
  library(readr)
  library(dplyr)
  library(ggplot2)
  library(knitr)
})
here::i_am("report/final_report.Rmd")
```

# Overview
This brief report presents a summary table of the top 10 countries by adolescent (10–19) HIV incidence in 2020 and a global trend figure.

```{r load-data, include=FALSE}
xlsx <- here("data","HIV_Epidemiology_Children_Adolescents_2021.xlsx")

# Try to detect correct header row automatically
df <- readxl::read_excel(
here::here("data","HIV_Epidemiology_Children_Adolescents_2021.xlsx"),
skip = 9,
.name_repair = "minimal"
)
# Keep first 5 cols and set names

df <- df[, 1:5, drop = FALSE]
names(df) <- c("country_region","year","age","indicator","value")

# Coerce types

df <- df |>
dplyr::mutate(
year = suppressWarnings(as.integer(year)),
value = suppressWarnings(readr::parse_number(as.character(value)))
)

# Filters used below

inc_name <- "Estimated incidence rate (new HIV infection per 1,000 uninfected population)"
ages <- "Age 10-19"
```

# Table: Top 10 countries (10–19), 2020

```{r table, echo=FALSE}
tbl <- df |>
  filter(age == ages, indicator == inc_name, year == 2020) |>
  group_by(country_region) |>
  summarise(incidence_2020 = mean(value, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(incidence_2020)) |>
  slice_head(n = 10)

dir.create(here("output","tables"), recursive = TRUE, showWarnings = FALSE)
readr::write_csv(tbl, here("output","tables","top10_adol_inc_2020.csv"))

kable(tbl, caption = "Top 10 countries by adolescent (10–19) HIV incidence in 2020")
```

# Figure: Global adolescent HIV incidence (10–19)

```{r figure, echo=FALSE}
trend <- df |>
  filter(age == ages, indicator == inc_name) |>
  group_by(year) |>
  summarise(incidence = mean(value, na.rm = TRUE), .groups = "drop") |>
  arrange(year)

dir.create(here("output","figures"), recursive = TRUE, showWarnings = FALSE)
p <- ggplot(trend, aes(year, incidence)) +
  geom_line() + geom_point() +
  labs(
    title = "Global adolescent HIV incidence (10–19)",
    x = "Year", y = "Incidence rate (per 1,000 uninfected)"
  ) +
  theme_minimal()
ggsave(here("output","figures","global_adol_inc_trend.png"), p, width = 8, height = 5, dpi = 300)

p
```
# Notes
- **Data source:** HIV_Epidemiology_Children_Adolescents_2021.xlsx  
- **Indicator:** Estimated incidence rate (new HIV infection per 1,000 uninfected population)  
- **Age group:** 10–19 years  
- **Author:** Akanshya Dash  
- **Date rendered:** `r Sys.Date()`

---
