---
title: "HIV Epidemiology – Children & Adolescents"
author: "Akanshya Dash"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(here); library(readxl); library(dplyr); library(readr)
  library(ggplot2); library(knitr)
})
here::i_am("report/final_report.Rmd")
dir.create(here("output","tables"),  recursive = TRUE, showWarnings = FALSE)
dir.create(here("output","figures"), recursive = TRUE, showWarnings = FALSE)
```

# Overview
This brief report presents a summary table of the top 10 countries by adolescent (10–19) HIV incidence in 2020 and a global trend figure.

```{r load-data, include=FALSE}
xlsx <- here("data","HIV_Epidemiology_Children_Adolescents_2021.xlsx")
inc_name <- "Estimated incidence rate (new HIV infection per 1,000 uninfected population)"
ages <- "Age 10-19"

looks_ok <- function(x){
if (!is.data.frame(x) || ncol(x) < 5) return(FALSE)
nms <- tolower(names(x))
any(grepl("country|region", nms)) &&
any(grepl("^year$", nms)) &&
any(grepl("^age", nms)) &&
any(grepl("indicator", nms)) &&
any(grepl("^value", nms))
}

df <- NULL
for (sh in readxl::excel_sheets(xlsx)) {
for (sk in 0:20) {
tmp <- try(suppressWarnings(read_excel(xlsx, sheet = sh, skip = sk, .name_repair="minimal")), silent=TRUE)
if (!inherits(tmp,"try-error") && looks_ok(tmp)) { df <- tmp; break }
}
if (!is.null(df)) break
}
stopifnot(!is.null(df))

# pick the first matching columns by NAME (not index)

nms <- names(df)
pick <- function(p) which(grepl(p, nms, ignore.case=TRUE))[1]
cols <- c(
pick("country|region"),
pick("^year$"),
pick("^age"),
pick("indicator"),
pick("^value")
)
df <- df[, cols, drop=FALSE]
names(df) <- c("country_region","year","age","indicator","value")

df <- df |>
mutate(
year = suppressWarnings(as.integer(year)),
value = suppressWarnings(readr::parse_number(as.character(value)))
)
```

# Table: Top 10 countries (10–19), 2020

```{r table, echo=FALSE}
tbl <- df |>
filter(age == ages, indicator == inc_name, year == 2020) |>
group_by(country_region) |>
summarise(incidence_2020 = mean(value, na.rm = TRUE), .groups = "drop") |>
arrange(desc(incidence_2020)) |>
slice_head(n = 10)

# save for the repo

readr::write_csv(tbl, here("output","tables","top10_adol_inc_2020.csv"))

kable(tbl, caption = "Top 10 countries by adolescent (10–19) HIV incidence in 2020")
```

# Figure: Global adolescent HIV incidence (10–19)

```{r figure, echo=FALSE}
trend <- df |>
filter(age == ages, indicator == inc_name) |>
group_by(year) |>
summarise(incidence = mean(value, na.rm = TRUE), .groups = "drop") |>
arrange(year)

p <- ggplot(trend, aes(year, incidence)) +
geom_line() + geom_point() +
labs(title = "Global adolescent HIV incidence (10–19)",
x = "Year", y = "Incidence per 1,000 uninfected") +
theme_minimal()

# save for the repo

ggsave(here("output","figures","global_adol_inc_trend.png"), p, width=8, height=5, dpi=300)

p
```

# Notes
- **Data source:** HIV_Epidemiology_Children_Adolescents_2021.xlsx  
- **Indicator:** Estimated incidence rate (new HIV infection per 1,000 uninfected population)  
- **Age group:** 10–19 years  
- **Author:** Akanshya Dash  
- **Date rendered:** `r Sys.Date()`
